Global code motion of congruent instructions

### 2017-09-16 cgo/paper.tex
final cgo

%% Journal information
%% Supplied to authors by publisher for camera-ready submission;
%% use defaults for review submission.
\acmJournal{PACMPL}
\acmVolume{1}
\acmNumber{CONF} % CONF = POPL or ICFP or OOPSLA
\acmArticle{1}
\acmYear{2017}
\acmMonth{1}
\acmDOI{} % \acmDOI{10.1145/nnnnnnn.nnnnnnn}
\startPage{1}

@@

%% Some recommended packages.
\usepackage{booktabs}   %% For formal tables:
                        %% http://ctan.org/pkg/booktabs
\usepackage{subcaption} %% For complex figures with subfigures/subcaptions
                        %% http://ctan.org/pkg/subcaption
\usepackage{comment}
\usepackage{amsmath}
\usepackage{hyperref}
\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{graphviz}
\usepackage{auto-pst-pdf}
\usepackage{etoolbox}
\usepackage{flushend}
\usepackage{needspace}
\usepackage{algpseudocode}
\usepackage[final]{pdfpages}

@@

%% Keywords
%% comma separated list
\keywords{keyword1, keyword2, keyword3}  %% \keywords are mandatory in final camera-ready submission

@@

\label{sec:intro}
Compiler techniques to remove redundant computations are composed of an analysis
phase that detects identical computations in the program and a transformation
phase that reduces the number of run-time computations.  Classical scalar
optimizations like Common Subexpression Elimination (CSE) \cite{dragonbook} work
very well on single basic-blocks (local level) but when it comes to detect
redundancies across basic-blocks (global level) these techniques fall short:
more complex passes like Global Common Subexpression Elimination (GCSE) and
Partial Redundancy Elimination (\PRE{}) have been designed to handle these cases
based on data-flow analysis \cite{morel1979global}.  At first these techniques
were described in the classical data-flow analysis framework, but later the use
of Static Single Assignment (\SSA{}) representation lowered their cost in terms
of compilation time \cite{briggs1994effective,chow1997new,kennedy1999partial}
and brought these techniques in the main stream: nowadays \SSA{} based \PRE{} is
available in industrial compilers like \GCC{} and \LLVM{}.

This paper describes \gcm{} (\GCM{}) of congruent instructions
\cite{briggs1997}, a technique that uses the information computed for \PRE{} to
detect identical computations but has a transformation phase whose goal differs
from \PRE{}: it removes identical computations from different branches of
execution.  These identical computations in different branches of execution are
not redundant computations at run-time and the number of run-time computations
is not reduced. It is not a redundancy elimination pass, and thus it has
different cost function and heuristics than \PRE{} or CSE. It is similar to
global code scheduling \cite{dragonbook,click1995global} in the sense that it
will only move computations. Code hoisting, for example, can reduce the critical
path length of execution in \ooo{} machines. As more instructions are available
at the hoisting point, the hardware has more independent instructions to
reorder. The following example illustrates how hoisting can improve performance
by exposing more ILP.

@@

Now if the computation of $tmax$ and $tmin$ are hoisted outside the
conditionals, the C code version would look like this:
\begin{verbatim}
float fun(float d, float min, float max, float a) {
  float inv = 1.0f / d;
  float x = (min - a) * inv;
  float y = (max - a) * inv;
  float tmin = x, tmax = y;
  if (inv < 0) { tmin = y; tmax = x; }
  return tmax + tmin;
}
\end{verbatim}

In this code the division operation can be executed in parallel with the two
subtractions because there are no dependencies among them. So the total number
of cycles will be $max(C_{div}, C_{sub}) + C_{mul} = C_{div} + C_{mul}$; since
$C_{div}$ is usually much greater than $C_{sub}$ \cite{x86,aarch64}. We have
seen the performance of an \ooo{} processor going up by $15\%$ on a benchmark
containing the above motivating example running in a tight loop.  In fact, there
are several advantages of \GCM{}:
\begin{itemize}
\item it helps reduce the code size of the program by replacing multiple
  instructions with one although the final code size may be higher because of
  improved inlining heuristics; functions become cheaper to inline when their
  code size decreases because the inliner heuristics depend on instruction
  count;
\item it exposes more instruction level parallelism to the later compiler
  passes. By hoisting identical computations to be executed earlier, instruction
  schedulers can move heavy computations earlier in order to avoid pipeline
  bubbles;
\item it reduces branch misprediction penalty on \ooo{} processors with
  speculative execution of branches: by hoisting or sinking expressions out of
  branches, it can effectively reduce the amount of code to be speculatively
  executed and hence reduce the critical path;
\item it reduces interference or register pressure with an appropriate
  cost-model: we have used a cost model in our implementation which only hoists
  or sinks when the register pressure is reduced;
\item it reduces the total number of instructions to be executed for SIMD
  architectures which execute all code in branches based on masking or
  predication;
\item it may improve loop vectorization by reducing a loop with control flow to
  a loop with a single basic-block, should all the instructions in a conditional get
  hoisted or sunk;
\item it enables more loop invariant code motion (LICM): as LICM passes, in
  general, cannot effectively reason about instructions within conditional
  branches in the context of loops, code-motion is needed to move instructions
  out of conditional expressions and expose them to LICM.
\end{itemize}

There has been a lot of work both in industry and academia to hoist and sink
code out of branches, and in general global scheduling
\cite{click1995global}. Some relate code-hoisting to code-size optimization
\cite{rosen1988global} and many \cite{barany2013, shobaki2013} use global
scheduling to improve performance. Most of the recent work on global scheduling
are done using integer linear programming (ILP,) which results in prohibitively
high compile time. To the best of our knowledge we have not found any reference
which explored \gcm{} of congruent instructions in as much detail us. A part of
our implementation (aggressive code hoisting) is already merged
in \LLVM{} trunk, however, a general implementation of \gcm{} of congruent (non
redundant) expressions is still missing from \GCC{} and \LLVM{} trunk. The main
contributions of this paper are:
\begin{itemize}
\item a new optimistic algorithm to move congruent instructions out of branches;
\item a cost model to reduce interference and hence, reduce spills;
\item a technique to maximize hoisting in an optimistic approach by partitioning
  the list of potential candidates sorted by their DFS visit number;
\item experimental evaluation of our implementation in \LLVM{} which combines
  \SSA{} based liveness analysis, and ranking of expressions to move very busy
  expressions in order to reduce register spills and improve performance in some
  cases.
\end{itemize}

@@

The algorithm for \gcm{} uses several common representations of the
program that we shortly describe below:
\begin{itemize}
\item Dominance (DOM) and Post-Dominance (PDOM) relations \cite{dragonbook} on a
  Control Flow Graph (\CFG{});
\item DJ-Graph \cite{Sreedhar1996} is a data structure that augments the
  dominator tree with join-edges to keep track of data-flow in a program. We use
  DJ-Graph to compute liveness of variables as illustrated in \cite{das2012};
\item Static Single Assignment (\SSA{}) \cite{cytron};
\item Global Value Numbering (\GVN{}) \cite{rosen1988global,click1995global}: to
  identify similar computations compilers use \GVN{}.  Each expression is given
  a unique number and the expressions that the compiler can prove to be
  identical are given the same number;
\item \MemorySSA{} \cite{novillo2007memory}: it is a factored use-def chain of
  memory operations that the compiler is unable to prove
  independent. \MemorySSA{} accelerates the access to the alias analysis
  information.
\end{itemize}

Dominance graphs, Global Value Numbers and Memory SSA are already available in
\LLVM{} but there is no facility to infer liveness of virtual registers. For
that we implemented DJ-Graph data structure that allows us to calculate
MergeSets which is used in the liveness analysis in our implementation of \GCM{}.

@@

\begin{verbatim}
// Return true if the merge-set of source node of a
// visited J-edge (incoming edge of lnode) is not
// the subset of the merge-set of lnode.
bool
still_inconsistent(Node lnode, JEdges JE,
                   Visited V, MergeSet MergeSet)
  for e in (all incoming edges to lnode):
    if e is in JE and e in V:
      if MergeSet(Source Node of e)
         is not a subset of MergeSet(lnode):
        return true
  return false
\end{verbatim}

Each basic block which has a join-edge (J-Edge) to its successor will have a
merge-set that is a subset of the merge-set of its successor \cite{das2005}. The
function \emph{still\_inconsistent} returns true if this invariant is not
satisfied which allows the function \emph{constructMergeSet} to repeat.

\begin{verbatim}
// Compute a merge-set breadth first order.
bool constructMergeSet_1(BFSList B, JEdges JE,
                         DomLevel DL)
  repeat = false
  V = empty list of visited edges
  for n in B:
    for e in (all incoming edges to n):
      if e is in JE and e not in V:
        V(e) = true
        let tmp = Source Node of e
        let tnode = Target Node of e
        let lnode = NULL
        while (DL(tmp) >= DL(tnode)):
          MergeSet(tmp) = MergeSet(tmp)
                          U MergeSet(tnode)
                          U {tnode}
          lnode = tmp
          tmp = dom-parent(tmp)
        repeat = still_inconsistent(lnode, JE, MergeSet)
  return repeat
\end{verbatim}
\newpage
\begin{verbatim}
// Construct merge-set of each node in
// Control-Flow-Graph G of a function.
void constructMergeSet(CFG G)
  B = Breadth First Order of G
  JE = JEdges of G
  DL = Path lengths (from root) of each node in G
  do: // Call until a fixed point is reached.
    Repeat = constructMergeSet_1(B, JE, DL)
  while (Repeat)
\end{verbatim}

@@

The original algorithm presented in \cite{das2012} has mismatched types in terms
of uses and nodes because each node can have many instructions and hence many
uses. Also, while iterating on the dominator of each user of a variable we may
reach to the beginning of a function, in that case the inner while loop needs to
terminate. These two cases were missing from the algorithm and we came across
them during implementation.

The \GCM{} pass can be broadly divided into the following steps that we will
describe in the rest of this section:
\begin{itemize}
\item find candidates (congruent instructions) suitable for \gcm{}
  (Section~\ref{subsec:finding-candidates}),
\item compute a point in the program where it is both legal
  (Section~\ref{subsec:legality}) and profitable
  (Section~\ref{subsec:cost-models}) to move the code,
\item move the code to hoist point or sink point
  (Section~\ref{subsec:optimistic}) and
\item update data structures to continue iterative \gcm{} (Section~\ref{subsec:optimistic}).
\end{itemize}

@@

\subsubsection{Legality of hoisting loads}
The availability of operand to the load (an address) is checked at the hoisting
point. If that is not available we try to rematerialize the address if
possible.  Along the path, from current position of the load instruction
backwards on the control flow to the hoisting point, we check whether there are
writes to memory that may alias with the load, in which case the candidate is
discarded. To iterate on the use-def chains for memory references the MemorySSA
infrastructure of \LLVM{} is used.

\subsubsection{Legality of hoisting stores}
For stores, we check the dependency requirements similar to the hoisting of
loads using the \MemorySSA{} of \LLVM{}. We check that the operands of the store
instruction are available at the hoisting point, and that there are no aliasing
loads or store along the path from the current position to the hoisting point.

@@

In order to be able to move an instruction that has side effects, we must
guarantee that the side-effects of that instruction before and after its
code-motion will produce the same side-effects.  For hoisting in particular, we
need to prove that on all execution paths from the hoisting point to the end of
the program the side effects appear in the same order. In general, we cannot
introduce a new computation along any path of execution without checking for
undefined behaviors. The algorithm \emph{compute\_downward\_safety} computes if
hoisting a set of instruction to a common hoisting point is downward-safe
\cite{steven1997advanced}. Since \GCM{} is only dealing with VBEs, computing
downward-safety ensures that no undefined behaviors are introduced as a result
of code-motion.
\newline
\begin{verbatim}
// Return true if a hoistable instruction is downward
// safe at hoist-point
bool compute_downward_safety(Node HP)
  Worklist = basic-blocks of hoistable instructions
  DT = The dominator tree
  HP = the hoist-point
  for each basic-block B in the DT starting at HP:
    if Worklist is empty
       return false // Path exists! not downward safe
    if B is in Worklist:
      remove B from worklist
      // Available at B => downward safe from B
      remove subtree with root at B
    if B is a leaf basic-block
      // Path exists! not downward safe
      return false
    if B dominates hoist-point
      return false  // Back edge
  return true
\end{verbatim}

\subsection{Profitability check (Cost models)}
\label{subsec:cost-models}
After the legality checks have passed, we check if a \gcm{} is profitable.  That
takes into account the impact \gcm{} would have on various parameters that might
affect runtime performance, e.g., impact on live-range, gain in the code
size. The current implementation makes effort to not regress in performance
across popular benchmarks and at the same time to reduce code size as much as
possible. Following are the cost models which are implemented:

@@

\begin{verbatim}
B0: b = m
    c = n
    if c is true then goto B3 else goto B4

B1: a0 = b<kill> + c<kill>

B2: a1 = b<kill> + c<kill>
\end{verbatim}

After hoisting a0 and a1 are removed and a copy of a0 as a01 is placed in B0
just before the branch. Since `b' and `c' are killed in `a0' and `a1', hoisting
the expressions will reduce the register pressure because two registers will be
freed.

Ideally, it should be okay to hoist all the instructions and a later
live-range-splitting \cite{cooper1998live} pass should make the right decision
of rematerializing the instruction should it be beneficial to do so. But the
current live-range splitting pass of \LLVM{} is not making the optimal decision
and we have found performance regressions while hoisting aggressively.

Moreover, \LLVM{} has a `getelementptr' instruction which computes the address
where a load or a store would happen. It is a scalar computation and gets
hoisted frequently even if the corresponding loads/stores would not be legal to
hoist.  In order to reduce the register pressure while hoisting loads, we have
restricted hoisting of address computations away from their corresponding loads
and stores when the loads and stores cannot be moved. This restriction is only
to mitigate the limitations of \LLVM{'}s register allocator and may be lifted in
the future, when the register allocation rematerialization pass has been
improved to catch these regressions.

There is a special case for hoisting load instructions when the hoist-point is
the predecessor basic-block for all the congruent loads. Even if the register
pressure would increase, we prefer to hoist loads in this case. The reason
being, that would make the loaded value available by the time it is used. Also,
because stores and calls are hoisted the least, see
Table~\ref{tab:code-motion-metric}, the performance does not change much whether
they are hoisted or not.

@@

In order to hoist a subset of identical instructions, we partition the list of
all candidates in a way to maximize the total number of hoist instructions.  By
sorting the list of all the candidates in the increasing order of their depth
first search discovery time stamp \cite{clrs} (DFSIn numbers,) we make sure that
candidates closer in the list have their common dominator nearby. In
Figure-\ref{fig:dfsin}, if B3, B4, B5, and B6 have identical computations and if
for some reason, they cannot be hoisted at B0, then we partition the set of
hoistable candidates in their DFSIn order. In this case the DFSIn ordering would
be B3, B4, B5, B6 which will allow the instructions in (B3, B4) to be hoisted at
B1 and those in (B5, B6) to be hoisted at B2. Even if instructions in (B3, B4)
could not be hoisted, for some reason, to B1, the instructions in (B5, B6) can
still be hoisted -- if legal and profitable -- to B2.

@@

\subsection{Code generation for the optimistic \gcm{} algorithm}
\label{subsec:optimistic}
Once all the legality and profitability checks are satisfied for a set of
congruent instructions, they are suitable candidates for hoisting or sinking. A
copy of the computation is inserted at the hoisting/sinking point along with any
instructions which needed to be rematerialized. Thereafter, all the computations
made redundant by the new copy are removed, and the \SSA{} form is restored by
updating the intermediate representation (IR) to reflect the changes. At the
same time \MemorySSA{} is also updated to get up-to-date information about
memory references.

@@

For hoisting, we find a common dominator dominating all a set of congruent
instructions and perform legality checks, as described in
Section~\ref{subsec:legality}. Often times it is not possible to hoist all the
instructions to one common dominator, due to legality constraints, e.g.,
intersecting side-effects, or profitability constraints, e.g., increasing
register pressure; in those cases, this algorithm would partition
(Section~\ref{subsec:partition}) the list of identical instructions into subsets
which can be partially hoisted to their respective common dominators.
\newpage
\begin{verbatim}
// Find hoistable candidates and do hoisting
int doCodeHoisting()
  For each value number VN with 2 or more instructions
    sort the instructions according to DFSIn numbers
    if first two I1, I2 are hoistable
        && downward-safe at hoist-point
        && profitable to be hoisted at hoist-point:
      continue the check for subsequent
        instructions with same VN
    else:
      proceed to check if I2 and I3 (if available)
      can be hoisted

@@

%\newpage
After no more candidates are hoistable, sinking is performed. Sinking is only
performed once because currently there are very few sinkable candidates
(Table~\ref{tab:code-motion-metric}) as SimplifyCFG of LLVM (which runs before
\GCM{}) has a code-sinking which already sinks several instructions.

@@

\section{Experimental Evaluation}
\label{sec:experimental-results}
% Experimental results: Register spills, compile time, run time. compile-time comparison w.r.t. gcc
For evaluation of \gcm{}, we built \SPEC{} with the patch. All the experiments
were conducted on an \xlinux{} machine and at -Ofast optimization level. Each
benchmark was run three times and the best result was taken. We collected
execution time and code-size which is listed in Table~\ref{tab:code-size}, other
compiler statistics are listed in Table~\ref{tab:spills}, Table~\ref{tab:stats}, and
Table~\ref{tab:code-motion-metric}.  We also measured compile time
Table~\ref{tab:compile-time} to see the impact of \gcm{}.

>>>>

%% Journal information
%% Supplied to authors by publisher for camera-ready submission;
%% use defaults for review submission.
%\acmJournal{PACMPL}
%\acmVolume{1}
%\acmNumber{CONF} % CONF = POPL or ICFP or OOPSLA
%\acmArticle{1}
%acmYear{2017}
%\acmMonth{1}
%\acmDOI{} % \acmDOI{10.1145/nnnnnnn.nnnnnnn}
\startPage{1}

%\setlength{\abovecaptionskip}{5pt plus 1pt minus 1pt} % Chosen fairly arbitrarily
%\setlength{\belowcaptionskip}{5pt plus 1pt minus 1pt} % Chosen fairly arbitrarily

@@

%% Some recommended packages.
\usepackage{booktabs}   %% For formal tables:
                        %% http://ctan.org/pkg/booktabs
\usepackage{subcaption} %% For complex figures with subfigures/subcaptions
                        %% http://ctan.org/pkg/subcaption
\usepackage{comment}
\usepackage{amsmath}
\usepackage{hyperref}
\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{graphviz}
\usepackage{auto-pst-pdf}
\usepackage{etoolbox}
\usepackage{flushend}
\usepackage{needspace}
\usepackage{algpseudocode}
\usepackage{enumitem}
\usepackage[final]{pdfpages}

@@

%% Keywords
%% comma separated list
\keywords{Optimizing Compilers, \GCC{}, \LLVM{}, Code Generation, Global Scheduling}  %% \keywords is optional

@@

\label{sec:intro}
Compiler techniques to remove redundant computations are composed of an analysis
phase that detects identical computations in the program and a transformation
phase that reduces the number of run-time computations.  Classical scalar
optimizations like Common Subexpression Elimination (CSE) \cite{dragonbook} work
very well on single basic-blocks (local level) but when it comes to detect
redundancies across basic-blocks (global level) these techniques fall short:
more complex passes like Global Common Subexpression Elimination (GCSE) and
Partial Redundancy Elimination (\PRE{}) have been designed to handle these cases
based on data-flow analysis \cite{morel1979global}.  At first these techniques
were described in the classical data-flow analysis framework, but later the use
of Static Single Assignment (\SSA{}) representation lowered their cost in terms
of compilation time \cite{briggs1994effective,chow1997new,kennedy1999partial}
and brought these techniques in the main stream: \SSA{} based \PRE{} is
available in many compilers like \GCC{} and \LLVM{}.

This paper describes \gcm{} (\GCM{}) of congruent instructions
\cite{briggs1997}, a technique that uses the information computed for \PRE{} to
detect identical computations but has a transformation phase whose goal differs
from \PRE{}: it removes identical computations from different branches of
execution.  These identical computations in different branches of execution are
not redundant computations at run-time and the number of run-time computations
is not reduced. It is not a redundancy elimination pass, and thus it has
different cost function and heuristics than \PRE{} or CSE. It is similar to
global code scheduling \cite{dragonbook,click1995global} in the sense that it
will only move computations. Code hoisting, for example, can reduce the critical
path length of execution in \ooo{} machines. As more instructions are available
at the hoisting point, the hardware has more independent instructions to
reorder. The following reduced example illustrates how hoisting can improve
performance by exposing more ILP.

@@

Now if the computation of $tmax$ and $tmin$ are hoisted outside the
conditionals, the C code version would look like:
\begin{verbatim}
float fun(float d, float min, float max, float a) {
  float inv = 1.0f / d;
  float x = (min - a) * inv;
  float y = (max - a) * inv;
  float tmin = x, tmax = y;
  if (inv < 0) { tmin = y; tmax = x; }
  return tmax + tmin;
}
\end{verbatim}

In this code the division operation can be executed in parallel with the two
subtractions because there are no dependencies among them. So the total number
of cycles will be $max(C_{div}, C_{sub}) + C_{mul} = C_{div} + C_{mul}$; since
$C_{div}$ is usually much greater than $C_{sub}$ \cite{x86,aarch64}. We have
seen the performance of the original benchmark (on an \ooo{} processor) improve
by $15\%$.  In fact, there are several advantages of \GCM{}:
\begin{itemize}[leftmargin=*,topsep=0pt]
\item it helps reduce the code size of the program by replacing multiple
  instructions with one although the final code size may be higher because of
  improved inlining heuristics; functions become cheaper to inline when their
  code size decreases because the inliner heuristics depend on instruction
  count;
\item it exposes more instruction level parallelism to the later compiler
  passes. By hoisting identical computations to be executed earlier, instruction
  schedulers can move heavy computations earlier in order to avoid pipeline
  bubbles;
\item it reduces branch misprediction penalty on \ooo{} processors with
  speculative execution of branches: by hoisting or sinking expressions out of
  branches, it can effectively reduce the amount of code to be speculatively
  executed and hence reduce the critical path;
\item it reduces interference or register pressure with an appropriate
  cost-model: we have used a cost model in our implementation which only hoists
  or sinks when the register pressure is reduced;
\item it reduces the total number of instructions to be executed for SIMD
  architectures which execute all code in branches based on masking or
  predication;
\item it may improve loop vectorization by reducing a loop with control flow to
  a loop with a single basic-block, should all the instructions in a conditional get
  hoisted or sunk;
\item it enables more loop invariant code motion (LICM): as LICM passes, in
  general, cannot effectively reason about instructions within conditional
  branches in the context of loops, code-motion is needed to move instructions
  out of conditional expressions and expose them to LICM.
\end{itemize}

There has been a lot of work both in industry and academia to hoist and sink
code out of branches, and in general global scheduling
\cite{click1995global}. Some relate code-hoisting to code-size optimization
\cite{rosen1988global} and many \cite{barany2013, shobaki2013} use global
scheduling to improve performance. Most of the recent work on global scheduling
are done using integer linear programming (ILP,) which results in prohibitively
high compile time. To the best of our knowledge we have not found any reference
which explored \gcm{} of congruent instructions in as much detail us. A part of
our implementation (aggressive code hoisting) is already merged
in \LLVM{} trunk, however, a general implementation of \gcm{} of congruent (non
redundant) expressions is still missing from \GCC{} and \LLVM{} trunk. The main
contributions of this paper are:
\begin{itemize}[leftmargin=*,topsep=0pt]
\item a new optimistic algorithm to move congruent instructions out of branches;
\item a cost model to reduce interference and hence, reduce spills;
\item a technique to maximize hoisting in an optimistic approach by partitioning
  the list of potential candidates sorted by their DFS visit number;
\item experimental evaluation of our implementation in \LLVM{} which combines
  \SSA{} based liveness analysis, and ranking of expressions to move very busy
  expressions in order to reduce register spills and improve performance in some
  cases.
\end{itemize}

@@

The algorithm for \GCM{} uses several common representations of the
program that we shortly describe below:
\begin{itemize}[leftmargin=*,topsep=0pt]
\item Dominance (DOM) and Post-Dominance (PDOM) relations \cite{dragonbook} on a
  Control Flow Graph (\CFG{});
\item DJ-Graph \cite{Sreedhar1996} is a data structure that augments the
  dominator tree with join-edges to keep track of data-flow in a program. We use
  DJ-Graph to compute liveness of variables as illustrated in \cite{das2012};
\item Static Single Assignment (\SSA{}) \cite{cytron};
\item Global Value Numbering (\GVN{}) \cite{rosen1988global,click1995global}: to
  identify similar computations compilers use \GVN{}.  Each expression is given
  a unique number and the expressions that the compiler can prove to be
  identical are given the same number;
\item \MemorySSA{} \cite{novillo2007memory}: it is a factored use-def chain of
  memory operations that the compiler is unable to prove
  independent. \MemorySSA{} accelerates the access to the alias analysis
  information.
\end{itemize}

Dominance graphs, Global Value Numbers and Memory SSA are already available in
\LLVM{} but there is no facility to infer liveness of virtual registers. So we
implemented DJ-Graph data structure that allowed us to calculate MergeSets which
was used in the liveness analysis during \GCM{}.

@@

\begin{verbatim}
// Return true if the merge-set of source node of a
// visited J-edge (incoming edge of lnode) is not
// the subset of the merge-set of lnode.
bool
stillInconsistent(Node lnode, JEdges JE,
                   Visited V, MergeSet MergeSet)
  for e in (all incoming edges to lnode):
    if e is in JE and e in V:
      if MergeSet(Source Node of e)
         is not a subset of MergeSet(lnode):
        return true
  return false
\end{verbatim}

Each basic block which has a join-edge (J-Edge) to its successor will have a
merge-set that is a subset of the merge-set of its successor \cite{das2005}. The
function \emph{stillInconsistent} returns true if this invariant is not
satisfied which allows the function \emph{constructMergeSet} to repeat.

\begin{verbatim}
// Compute a merge-set breadth first order.
bool constructMergeSet_1(BFSList B, JEdges JE,
                         DomLevel DL)
  repeat = false
  V = empty list of visited edges
  for n in B:
    for e in (all incoming edges to n):
      if e is in JE and e not in V:
        V(e) = true
        let tmp = Source Node of e
        let tnode = Target Node of e
        let lnode = NULL
        while (DL(tmp) >= DL(tnode)):
          MergeSet(tmp) = MergeSet(tmp)
                          U MergeSet(tnode)
                          U {tnode}
          lnode = tmp
          tmp = dom-parent(tmp)
        repeat = still_inconsistent(lnode, JE, MergeSet)
  return repeat
\end{verbatim}

\begin{verbatim}
// Construct merge-set of each node in
// Control-Flow-Graph G of a function.
void constructMergeSet(CFG G)
  B = Breadth First Order of G
  JE = JEdges of G
  DL = Path lengths (from root) of each node in G
  do: // Call until a fixed point is reached.
    Repeat = constructMergeSet_1(B, JE, DL)
  while (Repeat)
\end{verbatim}

@@

The original algorithm presented in \cite{das2012} has mismatched types in terms
of uses and nodes because each node can have many instructions and hence many
uses. Also, while iterating on the dominator of each user of a variable we may
reach to the beginning of a function, in that case the inner while loop needs to
terminate. These two cases were missing from the original algorithm.

The \GCM{} pass can be broadly divided into the following steps that we will
describe in the rest of this section:
\begin{itemize}[leftmargin=*,topsep=0pt]
\item find candidates (congruent instructions) suitable for \gcm{}
  (Section~\ref{subsec:finding-candidates}),
\item compute a point in the program where it is both legal
  (Section~\ref{subsec:legality}) and profitable
  (Section~\ref{subsec:cost-models}) to move the code,
\item move the code to hoist point or sink point
  (Section~\ref{subsec:optimistic}),
\item update data structures to continue iterative \gcm{} (Section~\ref{subsec:optimistic}).
\end{itemize}

@@

\subsubsection{Legality of hoisting loads}
The availability of operand to the load (an address) is checked at the hoisting
point. If that is not available we try to rematerialize the address if
possible.  Along the path, from current position of the load instruction
backwards on the control flow to the hoisting point, we check whether there are
writes to memory that may alias with the load, in which case the candidate is
discarded. To iterate on the use-def chains for memory references the MemorySSA
of \LLVM{} is used.

\subsubsection{Legality of hoisting stores}
For stores, we check the dependency requirements similar to the hoisting of
loads using the \MemorySSA{} of \LLVM{}. We check that the operands of the store
instruction are available at the hoisting point, and there are no aliasing
memory references along the hoisting path.

@@

In order to be able to move an instruction that has side effects, we must
guarantee that the side-effects of that instruction before and after its
code-motion will produce the same side-effects.  For hoisting in particular, we
need to prove that on all execution paths from the hoisting point to the end of
the program the side effects appear in the same order. In general, we cannot
introduce a new computation along any path of execution without checking for
undefined behaviors. The algorithm \emph{computeDownwardSafety} computes if
hoisting a set of instruction to a common hoisting point is downward-safe
\cite{steven1997advanced}. Since \GCM{} is only dealing with VBEs, computing
downward-safety ensures that no undefined behaviors are introduced as a result
of code-motion.
\newline
\begin{verbatim}
// Return true if a hoistable instruction is downward
// safe at hoist-point
bool computeDownwardSafety(Node HP)
  Worklist = basic-blocks of hoistable instructions
  DT = The dominator tree
  HP = the hoist-point
  for each basic-block B in the DT starting at HP:
    if Worklist is empty
       return false // Path exists! not downward safe
    if B is in Worklist:
      remove B from worklist
      // Available at B => downward safe from B
      remove subtree with root at B
    if B is a leaf basic-block
      // Path exists! not downward safe
      return false
    if B dominates hoist-point
      return false  // Back edge
  return true
\end{verbatim}

\subsection{Profitability check (Cost models)}
\label{subsec:cost-models}
After the legality checks have passed, we check if a \gcm{} is profitable.  That
takes into account the impact \gcm{} would have on various parameters that might
affect runtime performance, e.g., impact on live-range, gain in the code
size. The current implementation makes effort to not regress in performance
across popular benchmarks and at the same time to reduce code size as much as
possible. Following cost models are implemented:

@@

\begin{verbatim}
B0: b = m
    c = n
    if c is true then goto B3 else goto B4
B1: a0 = b<kill> + c<kill>
B2: a1 = b<kill> + c<kill>
\end{verbatim}

After hoisting, $a0$ and $a1$ are removed and a copy of $a0$ as $a1$ is placed
in $B0$ just before the branch. Since $b$ and $c$ are killed in $a0$ and $a1$,
hoisting the expressions will reduce the register pressure because two registers
will be freed.

Ideally, it should be okay to hoist all the instructions and a later
live-range-splitting \cite{cooper1998live} pass should make the right decision
but the current live-range splitting pass of \LLVM{} is not making the optimal
decision and we have found performance regressions while hoisting aggressively.

There is a special case for hoisting load instructions when the hoist-point is
the predecessor basic-block for all the congruent loads. Even if the register
pressure would increase, we prefer to hoist loads in this case. The reason
being, that would make the loaded value available by the time it is used.

@@

In order to hoist a subset of identical instructions, we partition the list of
all candidates in a way to maximize the total number of hoist instructions.  By
sorting the list of all the candidates in the increasing order of their depth
first search discovery time stamp \cite{clrs} (DFSIn numbers,) we make sure that
candidates closer in the list have their common dominator nearby. In
Figure-\ref{fig:dfsin}, if B3, B4, B5, and B6 have identical computations and if
for some reason, they cannot be hoisted at B0, then we partition the set of
hoistable candidates in their DFSIn order. In this case the DFSIn ordering would
be B3, B4, B5, B6 which will allow the instructions in (B3, B4) to be hoisted at
B1 and those in (B5, B6) to be hoisted at B2. Even if instructions in (B3, B4)
could not be hoisted, for some reason, to B1, the instructions in (B5, B6) can
still be hoisted - if legal and profitable - to B2.

@@

\subsection{Code generation for the optimistic \gcm{} algorithm}
\label{subsec:optimistic}
Once all the legality and profitability checks are satisfied for a set of
congruent instructions, they are suitable candidates for hoisting or sinking. A
copy of the computation is inserted at the hoisting/sinking point along with any
instructions which needed to be rematerialized. Thereafter, all the computations
made redundant by the new copy are removed, and the \SSA{} form is restored by
updating the intermediate representation (IR) to reflect the changes. At the
same time \MemorySSA{} is updated as well.

@@

For hoisting, we find a common dominator dominating all a set of congruent
instructions and perform legality checks, as described in
Section~\ref{subsec:legality}. Often times it is not possible to hoist all the
instructions to one common dominator, due to legality constraints, e.g.,
intersecting side-effects, or profitability constraints, e.g., increasing
register pressure; in those cases, this algorithm would partition
(Section~\ref{subsec:partition}) the list of identical instructions into subsets
which can be partially hoisted to their respective common dominators.

\begin{verbatim}
// Find hoistable candidates and do hoisting
int doCodeHoisting()
  For each value number VN with 2 or more instructions
    sort the instructions according to DFSIn numbers
    if first two I1, I2 are hoistable
        && downward-safe at hoist-point
        && profitable to be hoisted at hoist-point:
      continue the check for subsequent
        instructions with same VN
    else:
      proceed to check if I2 and I3 (if available)
      can be hoisted

@@

%\newpage
After no more candidates are hoistable, sinking is performed. Sinking is only
performed once because currently there are very few sinkable candidates
(Figure~\ref{tab:code-motion-metric}) as SimplifyCFG of LLVM (which runs before
\GCM{}) has a code-sinking which already sinks several instructions.

@@

\section{Experimental Evaluation}
\label{sec:experimental-results}
% Experimental results: Register spills, compile time, run time. compile-time comparison w.r.t. gcc
For evaluation of \gcm{}, we built \SPEC{} with the patch. All the experiments
were conducted on an \xlinux{} machine and at -Ofast optimization level. Each
benchmark was run three times and the best result was taken. We collected
execution time and code-size which is listed in Table~\ref{tab:code-size}, other
compiler statistics are listed in Table~\ref{tab:spills}, Table~\ref{tab:stats}, and
Figure~\ref{tab:code-motion-metric}.  We also measured compile time
Table~\ref{tab:compile-time} to see the impact of \gcm{}.

\begin{table}[h!]
  \begin{center}
    \scalebox{0.7}{
    \begin{tabular}{|l|c|c|c|}
      \hline
% TODO: Check if putting code-size or spills number is okay or only %age is required.
% llvm trunk vs. llvm trunk with code-motion
                &\% performance uplift & size reduction (Bytes) & size reduction (Bytes)\\
Benchmarks      & at -Ofast (high better)& at -Ofast (high better) & at -Os (high better)\\\hline
400.perlbench	& 1.00  & -9120  & 544  \\\hline
401.bzip2	& 1.00  & -1600  & 128  \\\hline
403.gcc	        & 1.00  & -58176 & 3384 \\\hline
429.mcf	        & 1.03  & 160    & 16   \\\hline
433.milc	& 1.00  & -656   & 224  \\\hline
444.namd	& 1.00  & -3864  & -1168\\\hline
445.gobmk	& 0.99  & -20480 & -264 \\\hline
447.dealII	& 0.95  & -27836 & 50934\\\hline
450.soplex	& 0.99  & 2220   & 348  \\\hline
453.povray	& 0.97  & 164    & -472 \\\hline
456.hmmer	& 0.99  & -3464  & -352 \\\hline
458.sjeng	& 1.02  & 440    & 1456 \\\hline
462.libquantum	& 1.00  & -96    & 0    \\\hline
464.h264ref	& 1.00  & -3424  & 864  \\\hline
470.lbm	        & 1.00  & 0      & 0    \\\hline
471.omnetpp	& 1.00  & 2688   & 200  \\\hline
473.astar	& 1.00  & -928   & 152  \\\hline
482.sphinx3	& 0.99  & 3008   & -176 \\\hline
483.xalancbmk	& 1.01  & 46492  & 4012 \\\hline
Geomean         & 0.997 &  & \\\hline
    \end{tabular}}
  \end{center}
  \caption{Execution time (ratio) and code size change (Bytes) with and without
    \GCM{} on \SPEC{}: performance uplift at in mcf and sjeng;
    code size improvement in dealII.}
  \label{tab:code-size}
\end{table}
